---
title: "Assess whether correlation tables are internally consistent, and reconstruct regression analyses from correlation matrices"
author: "Ian Hussey"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

IN DEVELOPMENT - NOT RECOMMENDED FOR USE
 
```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

# dependencies
library(lavaan)
library(janitor)
library(tidyverse)
library(knitr)
library(kableExtra)
library(Matrix)

```

# Simple way to calculate variance-covariance matricies and means vectors

```{r}

mtcars |>
  cor(use = "pairwise.complete.obs",
      method = "pearson") |>
  round(2)

a <- mtcars |>
  cov(use = "everything",
      method = "pearson")

cov2cor(a) |>
  round(2)

colMeans(mtcars) |>
  round(2)

```

# Correlation consistency

Given three correlations calculated within a given sample using use = "all", assess whether the third correlation is possible given the first two.

```{r}

correlation_consistency <- function(r_XY, r_XZ, r_YZ){
  # Since X, Y, and Z are z-scored vectors, the correlation coefficient between any pair is the cosine similarity.
  lower_bound <- cos(acos(r_XY) + acos(r_XZ))
  upper_bound <- cos(abs(acos(r_XY) - acos(r_XZ)))
  
  # correlations are bounded [-1, +1], so correct impossible results
  lower_bound <- ifelse(lower_bound < -1, -1, 
                        ifelse(lower_bound > +1, +1, lower_bound))
  
  upper_bound <- ifelse(upper_bound < -1, -1, 
                        ifelse(upper_bound > +1, +1, upper_bound))
  
  res <- data.frame(r_XY = r_XY,
                    r_XZ = r_XZ,
                    r_YZ = r_YZ,
                    r_YZ_lower_bound = lower_bound,
                    r_YZ_upper_bound = upper_bound) |>
    mutate(consistent = ifelse(r_YZ > r_YZ_lower_bound & r_YZ < r_YZ_upper_bound, TRUE, FALSE))
  
  return(res)
}

correlation_consistency(r_XY = .70, r_XZ = .80, r_YZ = .10)

```

# Convert a full cor matrix into a lower triangle (e.g., for reporting)

```{r}

cor_matrix_to_lower_triangle <- function(cor_matrix){
  cor_matrix[upper.tri(cor_matrix, diag = TRUE)] <- NA
  return(cor_matrix)
}

mat <- cor(mtcars)

cor_matrix_to_lower_triangle(mat) |>
  round(2)

```

# Convert an upper or lower triangle of correlations (e.g., as reported in an article) into a full cor matrix

```{r}

triangle_to_cor_matrix <- function(triangle){
  
  library(tibble)
  
  # if input is a df or tibble, convert to matrix
  if(is.data.frame(triangle) | is_tibble(triangle)){
    triangle <- as.matrix(triangle)
  } 
  if(!is.matrix(triangle)){
    stop("Input must be a matrix, data frame, or tibble")
  }
  
  # assess if the triangle is an upper triangle. I.e., are all the lower tri values NA and all the upper tri values non-NA? If so, transpose the matrix to make it a lower triangle. 
  if(all(is.na(triangle[lower.tri(triangle)])) &
     all(!is.na(triangle[upper.tri(triangle)]))){
    triangle <- t(triangle)
  }
  
  # create an empty matrix of the same dimension as 'triangle' to hold the mirrored values
  mirror_matrix <- matrix(0, nrow = nrow(triangle), ncol = ncol(triangle))
  
  # assign the lower triangle of 'mirror_matrix' with the lower triangle of 'triangle'
  mirror_matrix[lower.tri(mirror_matrix)] <- triangle[lower.tri(triangle)]
  
  # add the transposed 'mirror_matrix' to itself to complete the mirroring process
  full_matrix <- mirror_matrix + t(mirror_matrix) - diag(nrow(triangle))
  
  # fill diagonal with 1s
  diag(full_matrix) <- 1
  
  return(full_matrix)
}

# can handle lower triangles
mat <- cor_matrix_to_lower_triangle(cor(mtcars)) |> round(2)
mat
triangle_to_cor_matrix(mat)

# and upper triangles
mat <- t(cor_matrix_to_lower_triangle(cor(mtcars))) |> round(2)
mat
triangle_to_cor_matrix(mat)

```

# Positive definiteness

Is a correlation matrix of arbitrary size internally consistent?

Assumes no missingness/listwise deletions, and missingness and deletion method may or may not have been fully reported.

```{r}

positive_definite <- function(cor_matrix){
  all(eigen(cor_matrix)$values > 0)
}

mat <- cor(mtcars)
positive_definite(mat)

```

# Inspect distribution of eigenvalues

```{r}

dat <- data.frame(eigenvalues = eigen(mat)$values)

ggplot(dat, aes(eigenvalues)) +
  geom_histogram(closed = "right", boundary = 0) +
  geom_vline(xintercept = 0, linetype = "dashed")

```

# Range constraints

Is it worth simply checking that all correlations are [-1, 1]? Seems obvious, but maybe some violate this. Ask frank?

# Find nearest positive definite matrix

```{r}

correlation <- 
  lavaan::getCov(x = c(
    1.00,  0.32,  0.34, -0.09,  0.21, -0.90, # NB it is only this -.90 value that was changed from the original .30, a change of -1.20
           1.00,  0.45, -0.54,  0.54,  0.64,
                  1.00, -0.26,  0.32,  0.34, 
                         1.00, -0.52, -0.49, 
                                1.00,  0.50,
                                       1.00
  ),
  lower = FALSE,
  diagonal = TRUE, 
  )

# assess if PD
positive_definite(correlation)

# new PD matrix
nearest_positive_definite <- function(cor_matrix){
  Matrix::nearPD(cor_matrix, corr = TRUE)$mat
}

corrected_correlation <- nearest_positive_definite(correlation)
 
# assess if PD
positive_definite(corrected_correlation)

# print
corrected_correlation |>
  round_half_up(2)

# diff between original and corrected
correlation - nearest_positive_definite(correlation) |>
  round_half_up(2) |>
  cor_matrix_to_lower_triangle() |>
  t()

```

# PD violations

```{r}

library(MASS)

variables <- 5
rho <- 0.50
participants <- 100

mean_vector <- rep(0, variables)
cov_matrix <- matrix(rep(rho, variables * variables), nrow = variables, ncol = variables)
diag(cov_matrix) <- 1

dat_sim <- 
  mvrnorm(participants, mean_vector, cov_matrix) |>
  as.data.frame()

mat <- cor(dat_sim)
positive_definite(mat)

```


```{r}

# uniform distribution
#lower_tri <- matrix(runif((variables*(variables-1))/2, min = -1, max = 1), ncol = variables)

simulate_cor_matrix_norm <- function(n_variables, population_r_mean, population_r_sd){
  
  # draw correlations from a normal distribution fitting a lower triangle of a matrix
  lower_tri <- matrix(rnorm((n_variables*(n_variables-1))/2, 
                            mean = population_r_mean, 
                            sd = population_r_sd), 
                      ncol = n_variables)
  
  # correct correlations to [-1, 1] bounds
  lower_tri <- ifelse(lower_tri > 1, 1,
                      ifelse(lower_tri < -1, -1, lower_tri))
  
  # generate a correlation matrix from this lower triangle
  raw_cor_matrix <- lavaan::getCov(x = lower_tri, lower = TRUE, diagonal = FALSE)
  
  # create the nearest positive definite version of this matrix
  pd_cor_matrix <- nearPD(raw_cor_matrix)$mat
  
  # calculate differences between the two matrices
  diff_matrix <- pd_cor_matrix - raw_cor_matrix

  # results
  res <- list(
    raw_cor_matrix = raw_cor_matrix,
    pd_cor_matrix  = pd_cor_matrix,
    diff_matrix    = diff_matrix,
    r_mean         = mean(cor_matrix_to_lower_triangle(pd_cor_matrix), na.rm = TRUE),
    r_sd           = sd(cor_matrix_to_lower_triangle(pd_cor_matrix), na.rm = TRUE),
    diff_r_mean    = mean(cor_matrix_to_lower_triangle(diff_matrix), na.rm = TRUE),
    diff_r_sd      = sd(cor_matrix_to_lower_triangle(diff_matrix), na.rm = TRUE),
    max_abs_diff_r = max(abs(cor_matrix_to_lower_triangle(diff_matrix)), na.rm = TRUE)
  )
  
  res_rounded <- lapply(res, function(x) janitor::round_half_up(x, digits = 3))
  
  return(res_rounded)
}

#     cov_matrix_is_pd = positive_definite(cor(dat_sim)),

# dat_sim <- 
#   mvrnorm(n     = n_participants, 
#           mu    = rep(0, variables), 
#           Sigma = pd_matrix) |>
#   as.data.frame()


simulate_cor_matrix_norm(n_variables = 5, 
                         population_r_mean = 0.3, 
                         population_r_sd = 0.2)


```

```{r}

simulate_cor_matrix_uni <- function(n_variables, population_r_min, population_r_max){

  # draw correlations from a uniform distribution fitting a lower triangle of a matrix
  lower_tri <- matrix(runif((n_variables*(n_variables-1))/2, 
                            min = population_r_min, 
                            max = population_r_max),
                      ncol = n_variables)
  
  # correct correlations to [-1, 1] bounds
  lower_tri <- ifelse(lower_tri > 1, 1,
                      ifelse(lower_tri < -1, -1, lower_tri))
  
  # generate a correlation matrix from this lower triangle
  raw_cor_matrix <- lavaan::getCov(x = lower_tri, 
                                   lower = TRUE, 
                                   diagonal = FALSE,
                                   names = paste0("X", seq(from = 1, to = n_variables)))
  
  # create the nearest positive definite version of this matrix
  pd_cor_matrix <- nearPD(raw_cor_matrix)$mat
  
  # calculate differences between the two matrices
  diff_matrix <- pd_cor_matrix - raw_cor_matrix

  # results
  res <- list(
    raw_cor_matrix = raw_cor_matrix,
    pd_cor_matrix  = pd_cor_matrix,
    diff_matrix    = diff_matrix,
    r_mean         = mean(cor_matrix_to_lower_triangle(pd_cor_matrix), na.rm = TRUE),
    r_sd           = sd(cor_matrix_to_lower_triangle(pd_cor_matrix), na.rm = TRUE),
    diff_r_mean    = mean(cor_matrix_to_lower_triangle(diff_matrix), na.rm = TRUE),
    diff_r_sd      = sd(cor_matrix_to_lower_triangle(diff_matrix), na.rm = TRUE),
    max_abs_diff_r = max(abs(cor_matrix_to_lower_triangle(diff_matrix)), na.rm = TRUE)
  )
  
  res_rounded <- lapply(res, function(x) janitor::round_half_up(x, digits = 3))
  
  return(res_rounded)
}

res <- simulate_cor_matrix_uni(n_variables = 6, 
                               population_r_min = -0.20, 
                               population_r_max = +0.70)

#     cov_matrix_is_pd = positive_definite(cor(dat_sim)),

# dat_sim <-
#   mvrnorm(n     = 100,
#           mu    = rep(0, 5),
#           Sigma = res$pd_cor_matrix,
#           empirical = TRUE) |>
#   as.data.frame()

```


```{r}

# TODO need to add bounding check to disturb_pd_cor_matrix()

res <- simulate_cor_matrix_uni(n_variables = 5, 
                               population_r_min = -0.20, 
                               population_r_max = +0.70)

disturb_pd_cor_matrix <- function(pd_cor_matrix, item_to_offset, offset){
  cor_matrix <- pd_cor_matrix
  cor_matrix[item_to_offset+1, 1] <- cor_matrix[item_to_offset+1, 1] + offset
  cor_matrix[1, item_to_offset+1] <- cor_matrix[1, item_to_offset+1] + offset
  return(cor_matrix)
}

disturb_pd_cor_matrix(res$pd_cor_matrix, item_to_offset = 1, offset = 0.10) |>
  positive_definite()

```

proportion of pd raw matrices

```{r}

sim_uni <- function(n_variables, population_r_min, population_r_max){
  res <- simulate_cor_matrix_uni(n_variables = n_variables, 
                                 population_r_min = population_r_min, 
                                 population_r_max = population_r_max) 
  results <- positive_definite(res$raw_cor_matrix)
  return(results)
}

sim_results1 <- replicate(n = 10000, expr = sim_uni(n_variables = 5, population_r_min = -0.00, population_r_max = +0.50)) |> mean()
sim_results2 <- replicate(n = 10000, expr = sim_uni(n_variables = 5, population_r_min = -0.20, population_r_max = +0.70)) |> mean()
sim_results3 <- replicate(n = 10000, expr = sim_uni(n_variables = 5, population_r_min = -0.40, population_r_max = +0.90)) |> mean()
sim_results4 <- replicate(n = 10000, expr = sim_uni(n_variables = 5, population_r_min = -0.25, population_r_max = +0.25)) |> mean()
sim_results5 <- replicate(n = 10000, expr = sim_uni(n_variables = 5, population_r_min = -0.45, population_r_max = +0.45)) |> mean()
sim_results6 <- replicate(n = 10000, expr = sim_uni(n_variables = 5, population_r_min = -0.65, population_r_max = +0.65)) |> mean()

sim_results1
sim_results2
sim_results3
sim_results4
sim_results5
sim_results6

sim_results7 <- replicate(n = 1, expr = sim_uni(n_variables = 6, population_r_min = -0.00, population_r_max = +0.50)) |> mean()
sim_results8 <- replicate(n = 10000, expr = sim_uni(n_variables = 20, population_r_min = -0.20, population_r_max = +0.70)) |> mean()
sim_results9 <- replicate(n = 10000, expr = sim_uni(n_variables = 20, population_r_min = -0.40, population_r_max = +0.90)) |> mean()

sim_results7
sim_results8
sim_results9

```

- nonPDness is most detectable when both the individual correlations and variance among correlations are high. Even then, 

```{r}

sim_norm <- function(n_variables, population_r_mean, population_r_sd){
  res <- simulate_cor_matrix_norm(n_variables = n_variables, 
                                  population_r_mean = population_r_mean, 
                                  population_r_sd = population_r_sd) 
  results <- positive_definite(res$raw_cor_matrix)
  return(results)
}

sim_results1n <- replicate(n = 10000, expr = sim_norm(n_variables = 5, population_r_mean = 0.25, population_r_sd = +0.10)) |> mean()
sim_results2n <- replicate(n = 10000, expr = sim_norm(n_variables = 5, population_r_mean = 0.25, population_r_sd = +0.20)) |> mean()
sim_results3n <- replicate(n = 10000, expr = sim_norm(n_variables = 5, population_r_mean = 0.25, population_r_sd = +0.30)) |> mean()
sim_results4n <- replicate(n = 10000, expr = sim_norm(n_variables = 5, population_r_mean = 0.00, population_r_sd = +0.10)) |> mean()
sim_results5n <- replicate(n = 10000, expr = sim_norm(n_variables = 5, population_r_mean = 0.00, population_r_sd = +0.20)) |> mean()
sim_results6n <- replicate(n = 10000, expr = sim_norm(n_variables = 5, population_r_mean = 0.00, population_r_sd = +0.30)) |> mean()

sim_results1n
sim_results2n
sim_results3n
sim_results4n
sim_results5n
sim_results6n

```


proportion of PD disturbed matrices

```{r}

sim <- function(n_variables, population_r_min, population_r_max, item_to_offset, offset){
  res <- simulate_cor_matrix_uni(n_variables = n_variables, 
                                 population_r_min = population_r_min, 
                                 population_r_max = population_r_max) 
  results <- disturb_pd_cor_matrix(res$pd_cor_matrix, item_to_offset = item_to_offset, offset = offset) |>
    positive_definite()
  return(results)
}

sim_results <- 
  replicate(n = 10000, 
            expr = sim(n_variables = 5,
                       population_r_min = -0.20, 
                       population_r_max = +0.70,
                       item_to_offset = 1,
                       offset = +0.20)) |>
  mean()

sim_results

```

```{r}

sim2 <- function(n_variables, population_r_min, population_r_max, item_to_offset, offset){
  res <- simulate_cor_matrix_uni(n_variables = n_variables, 
                                 population_r_min = population_r_min, 
                                 population_r_max = population_r_max) 
  results <- 
    disturb_pd_cor_matrix(res$pd_cor_matrix, item_to_offset = item_to_offset, offset = offset) |>
    disturb_pd_cor_matrix(item_to_offset = item_to_offset + 1, offset = offset) |>
    positive_definite()
  return(results)
}

sim_results2 <- 
  replicate(n = 10000, 
            expr = sim2(n_variables = 5,
                        population_r_min = -0.20, 
                        population_r_max = +0.70,
                        item_to_offset = 1,
                        offset = +0.20)) |>
  mean()

sim_results2

```

```{r}

sim3 <- function(n_variables, population_r_min, population_r_max, item_to_offset, offset){
  res <- simulate_cor_matrix_uni(n_variables = n_variables, 
                                 population_r_min = population_r_min, 
                                 population_r_max = population_r_max) 
  results <- 
    disturb_pd_cor_matrix(res$pd_cor_matrix, item_to_offset = item_to_offset, offset = offset) |>
    disturb_pd_cor_matrix(item_to_offset = item_to_offset + 1, offset = offset) |>
    disturb_pd_cor_matrix(item_to_offset = item_to_offset + 2, offset = offset) |>
    positive_definite()
  return(results)
}

sim_results3 <- 
  replicate(n = 10000, 
            expr = sim3(n_variables = 5,
                        population_r_min = -0.20, 
                        population_r_max = +0.70,
                        item_to_offset = 1,
                        offset = +0.20)) |>
  mean()

sim_results3

```

```{r}

sim4 <- function(n_variables, population_r_min, population_r_max, item_to_offset, offset){
  res <- simulate_cor_matrix_uni(n_variables = n_variables, 
                                 population_r_min = population_r_min, 
                                 population_r_max = population_r_max) 
  results <- 
    disturb_pd_cor_matrix(res$pd_cor_matrix, item_to_offset = item_to_offset, offset = offset) |>
    disturb_pd_cor_matrix(item_to_offset = item_to_offset + 1, offset = offset*-1) |>
    positive_definite()
  return(results)
}

sim_results4 <- 
  replicate(n = 10000, 
            expr = sim4(n_variables = 5,
                        population_r_min = -0.20, 
                        population_r_max = +0.70,
                        item_to_offset = 1,
                        offset = +0.20)) |>
  mean()

sim_results4

```








